### 2023-08-21 proofs/main.tex
WIP: leftovers
WIP: description, notation

@@

\subsubsection{Variables and Types}
In the algorithmic system, we extend the grammar of types
by adding positive and negative \emph{algorithmic variables}
($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
They represent the unknown types, which will be inferred by the algorithm.
This way, we add two base cases to the grammar of 
positive and negative types, and use highlight to denote that the type
can potentially contain algorithmic variables.

@@

\subsubsection{Contexts}

Unannotated set of algorithmic variables is denoted as $[[Ξ]]$,
and it is used in the anti-unification algorithm.
$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ are used to denote
that the algorithmic type is well-formed in the contexts
$[[Γ]]$ and $[[Ξ]]$, which means that each algorithmic variable
of the type is contained in $[[Ξ]]$, and each free non-algorithmic type variable
of the type is contained in $[[Γ]]$.

Typically each algorithmic variable is associated with a
context it must be instantiated in 
(i.e. the context in which the type replacing the variable must be well-formed).
This association is represented by an \emph{algorithmic variable context} $[[Θ]]$.
\begin{definition}[Algorithmic Context]
  \hfill \\
  Algorithmic Context $[[Θ]]$ is represented by a set of 
  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
  and $[[Γ]]$ is a context in which they must be instantiated.
  We assume that no two entries associating the same variable
  appear in $[[Θ]]$.

@@

\subsubsection{Subsitutions}

@@

@@

\begin{definition}[Monadic Substitution Composition]
  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|{Γ}]]$.
\end{definition}

>>>>

\subsection{Equalities}
For simplicity, assume alpha-equivalent terms equal. 
This way, we assume that substitutions do not capture bound variables.
Besides, we equate
$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
and lift these equations transitively and congruently 
to the whole system.

\subsection{Contexts}

\begin{definition}[Declarative Type Context]
  \hfill \\
  Declarative type context $[[Γ]]$ is represented by a set of 
  type variables. The concatenation $[[Γ1, Γ2]]$ means the 
  union of two contexts  $[[{Γ1} ∪ {Γ2}]]$.
\end{definition}

\subsection{Substitutions}

\begin{definition}[Monadic Substitution Composition]
  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|{Γ}]]$.
\end{definition}
Notice that the result of $[[σ2 <=< σ1]]$ depends on the 
specification of $[[σ1]]$, which is not unique. 
However, we assume that the used specification clear from the 
context of the proof.

Sometimes it is convenient to construct substitution 
explicitly mapping each variable from a list (or a set)
to a type. Such substitutions are denoted as $[[iPs / pas]]$
and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
the corresponding types.
\begin{definition}[Explicit Substitution]
  \hfill
  \begin{itemize}
    \item [$-$]
      Suppose that $[[nas]]$ is a list of negative type variables,
      and $[[iNs]]$ is a list of negative types of the same length 
      Then $[[iNs / nas]]$ denotes a substitution such that 
      \begin{enumerate}
        \item for $[[αi⁺ ∊ {nas}]]$, $[[ [iNs / nas] αi⁺]] = [[iNi]]$;
        \item for $[[β⁺ ∉ {nas}]]$, $[[ [iNs / nas] β⁺]] = [[β⁺]]$.
      \end{enumerate}
    \item [$+$]
      Positive explicit substitution $[[iPs / pas]]$
      is defined symmetrically.
  \end{itemize}
\end{definition}

@@

In the algorithmic system, we extend the grammar of types
by adding positive and negative \emph{algorithmic variables}
($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
They represent the unknown types, which will be inferred by the algorithm.
This way, we add two base cases to the grammar of 
positive and negative types, and use highlight to denote that the type
can potentially contain algorithmic variables.

@@

\subsection{Algorithmic Renaming}

In several places of our algorithm, 
we turn a declarative type into the algorithmic one
via replacing certain type variables with fresh algorithmic variables.
This is done via algorithmic renaming

\begin{definition}
  Suppose that $[[nas]]$ is a list of negative type variables,

\subsection{Contexts}

\begin{definition}[Algorithmic Type Context]
  \hfill \\
  Algorithmic type context $[[Ξ]]$ is represented by a set of 
  \emph{algorithmic} type variables ($[[α̂⁺]]$, $[[α̂⁻]]$, $[[β̂⁺]]$, \dots).
\end{definition}

$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ are used to denote
that the algorithmic type is well-formed in the contexts
$[[Γ]]$ and $[[Ξ]]$, which means that each algorithmic variable
of the type is contained in $[[Ξ]]$, and each free declarative type variable
of the type is contained in $[[Γ]]$.

Algorithmic Type Context are used in the unification algorithm.
In the subtyping algorithm, 
the context needs to remember additional information.
In the subtyping context, each algorithmic variable is associated with a
context it must be instantiated in 
(i.e. the context in which the type replacing the variable must be well-formed).
This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
\begin{definition}[Algorithmic Subtyping Context]
  \hfill \\
  Algorithmic Subtyping Context $[[Θ]]$ is represented by a set of 
  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
  and $[[Γ]]$ is a context in which they must be instantiated.
  We assume that no two entries associating the same variable
  appear in $[[Θ]]$.

@@

\subsection{Subsitutions}

@@

We also define normalization of a substitution pointwise:
\begin{definition}[Substitution Normalization]
  For a substitution $[[σ]]$, we define $[[nf(σ)]]$
  as a substitution that maps $[[α±]]$ into $[[nf([σ]α±)]]$.
\end{definition}

@@

### 2023-08-21 proofs/tex/alg-neg-subtyping-lemmas.tex
WIP: description, notation

\begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$ and 
        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
        $[[Θ ⊢ SC]]$ and 
        for any normalized $[[uσ]]$ such that $[[Θ ⊢ uσ : SC]]$,
        $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.

    Suppose that $[[uσ]]$ is normalized and $[[Θ ⊢ uσ : SC]]$,
    Let us consider the last rule to infer this judgment. 
    \begin{caseof}
        \item \ruleref{\ottdruleAArrowLabel}. Then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ has shape
        $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ SC]]$\\
        On the next step, the the algorithm makes two recursive calls:
        $[[G;Θ ⊨ uP ≥ iQ ⫤ SC1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ SC2]]$
        and returns $[[Θ ⊢ SC1 & SC2 = SC]]$ as the result.

        By the soundness of constraint merge (\cref{lemma:merge-soundness}),
        $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$.
        Then by the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}), $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$; 
        and by the induction hypothesis, $[[ Γ ⊢ [uσ]uN' ≤ iM' ]]$.
        This way, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ ⊢ [uσ](uP → uN') ≤ iQ → iM']]$.

@@

\begin{lemma}[Completeness of the Negative Subtyping]
    \label{lemma:neg-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, such that $[[Θ ⊢ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    \begin{caseof}
        \item $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
        Then $[[ uN ]] = [[ ↑uP ]]$, since
        the substitution $[[ [uσ]uN ]]$ must preserve the 
        top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [uσ]uN ≈ iM ]]$.
        The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
         \cref{lemma:pos-subt-completeness}: notice that the algorithm does not make a recursive call, 
        and the difference in the induction statement for the positive and 
        the negative case here does not matter.

@@

        Notice that from the inversion of $[[Γ ⊢ iM]]$, we have: $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM']]$;
        from the inversion of $[[Γ ; Θ ⊢ uN]]$, we have: $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ; Θ ⊢ uN']]$;
        and since $[[uN]]$ does not contain negative unification variables,
        $[[uN']]$ does not contain negative unification variables either.

        This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
        obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2]]$ and $[[Θ ⊢ uσ : SC2]]$.
        Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ uσ : SC1]]$.

        Finally, we need to show that the merge of $[[SC1]]$ and $[[SC2]]$ is successful and
        satisfies the required properties.
        To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness}).
        This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, $[[Θ ⊢ SC]]$, and $[[Θ ⊢ uσ : SC]]$, as required.

>>>>

\begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
        $[[Θ ⊢ SC]]$ and 
        for any normalized $[[uσ]]$ such that $[[(Θ ⊢ SC) ⊢ uσ]]$,
        $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.

    Suppose that $[[uσ]]$ is normalized and $[[(Θ ⊢ SC) ⊢ uσ]]$,
    Let us consider the last rule to infer this judgment. 
    \begin{caseof}
        \item \ruleref{\ottdruleAArrowLabel}. Then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ has shape
        $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ SC]]$\\
        On the next step, the the algorithm makes two recursive calls:
        $[[G;Θ ⊨ uP ≥ iQ ⫤ SC1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ SC2]]$
        and returns $[[Θ ⊢ SC1 & SC2 = SC]]$ as the result.

        By the soundness of constraint merge (\cref{lemma:merge-soundness}),
        $[[(Θ ⊢ SC1) ⊢ uσ]]$ and $[[(Θ ⊢ SC2) ⊢ uσ]]$.
        Then by the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}), $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$; 
        and by the induction hypothesis, $[[ Γ ⊢ [uσ]uN' ≤ iM' ]]$.
        This way, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ ⊢ [uσ](uP → uN') ≤ iQ → iM']]$.

@@

\begin{lemma}[Completeness of the Negative Subtyping]
    \label{lemma:neg-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, such that $[[Θ ⊢ SC]]$ and moreover, $[[(Θ  ⊢  SC) ⊢ uσ]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    \begin{caseof}
        \item $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
        Then $[[ uN ]] = [[ ↑uP ]]$, since
        the substitution $[[ [uσ]uN ]]$ must preserve the 
        top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [uσ]uN ≈ iM ]]$.
        The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
         \cref{lemma:pos-subt-completeness}: notice that the algorithm does not make a recursive call, 
        and the difference in the induction statement for the positive and 
        the negative case here does not matter.

@@

        Notice that from the inversion of $[[Γ ⊢ iM]]$, we have: $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM']]$;
        from the inversion of $[[Γ ; dom(Θ) ⊢ uN]]$, we have: $[[Γ ; dom( Θ) ⊢  uP]]$ and $[[Γ ; dom( Θ) ⊢  uN']]$;
        and since $[[uN]]$ does not contain negative unification variables,
        $[[uN']]$ does not contain negative unification variables either.

        This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
        obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2]]$ and $[[(Θ  ⊢  SC2) ⊢ uσ]]$.
        Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1]]$ and $[[(Θ  ⊢  SC1) ⊢ uσ]]$.

        Finally, we need to show that the merge of $[[SC1]]$ and $[[SC2]]$ is successful and
        satisfies the required properties.
        To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness}).
        This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, $[[Θ ⊢ SC]]$, and $[[(Θ  ⊢  SC) ⊢ uσ]]$, as required.
